Manuscript Working Draft

### 2023-04-20 manuscript.qmd
cli commit 3

**Distinguishing developmental change from experience effects**. One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development. To this point, cognitive abilities (e.g, verbal ability, problem-solving) normatively grow earlier in development and often decline in late life (e.g., memory, speed of processing). However, the observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010a] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a]. Some designs (specifically, accelerated longitudinal designs) are especially well suited for discovering these effects and modeling them. While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design. Moreover, it is possible that in later waves, patterns of longitudinal missing data will allow some analyses to assess the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

@@

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021; xxxxx et al. (xxxx). Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

@@

## Linear vs non-linear models
Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used. Repeated assessments can be based on either continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016a]. For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, d) linear growth models for a count outcome assuming a zero-inflated Poisson model and e) discrete- and continuous-time survival analysis for a binary outcome. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data. These models can account for issues that can occur when working with discrete outcomes, including overdispersion (when the variance is higher than would be expected based on a given distribution) and zero-inflation [when more zeros occur than is expected based on a given distribution; see @lenz2016]. Models involving zero-inflation parameters are referred to as two-part models, given one part of the model predicts the zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of two-part models for longitudinal data]. Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014]. Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous indicators [e.g., @little2013]. However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].

@@

>>>>

**Distinguishing developmental change from experience effects**. One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development. To this point, cognitive abilities (e.g, verbal ability, problem-solving) normatively grow earlier in development and often decline in late life (e.g., memory, speed of processing). However, the observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a]. Some designs (specifically, accelerated longitudinal designs) are especially well suited for discovering these effects and modeling them. While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design. Moreover, it is possible that in later waves, patterns of longitudinal missing data will allow some analyses to assess the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

@@

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield20210603. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021; xxxxx et al. (xxxx). Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].

@@

## Linear vs non-linear models
Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used. Repeated assessments can be based on either continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016]. For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, d) linear growth models for a count outcome assuming a zero-inflated Poisson model and e) discrete- and continuous-time survival analysis for a binary outcome. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data. These models can account for issues that can occur when working with discrete outcomes, including overdispersion (when the variance is higher than would be expected based on a given distribution) and zero-inflation [when more zeros occur than is expected based on a given distribution; see @lenz2016]. Models involving zero-inflation parameters are referred to as two-part models, given one part of the model predicts the zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of two-part models for longitudinal data]. Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014a]. Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous indicators [e.g., @little2013]. However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].

@@

---

### References

::: {#refs}
:::
